### 2022-11-08 Manuscripts/deNovoRates.tex
modified:   Manuscripts/deNovoRates.tex
modified:   Modeling/DeNovoProb.jl

For all calculations we used a GC-content of 0.42 which is reasonably close to the global GC-content of human, \textit{Drosophila melanogaster}, and \textit{Saccharomyces cerevisiae} genomes. We used a mutation rate of $7.8\times10^{-9}$ mutations per nucleotide position per generation, which corresponds to spontaneous mutation rate estimated in a \textit{Drosophila melanogaster} population that was subjected to periodic population bottlenecks \cite[mutation accumulation line,][]{drosophilamutrate}. We used mutation bias data from the same study \citep{drosophilamutrate} which was in agreement with another previous study on human pseudogenes \citep{humanmutrate}. shows the values of different mutation rates.

@@

If a gene is found in one species but not in sister taxa, then the most common assumption is that this gene emerged only in one specific species. We asked what is the chance that the gene was present in a common ancestor, but was lost in all lineages except one. To this end we calculated the probability of gene emergence and gene loss (Materials and Methods). Specifically, the probability of gene gain can be defined as the sum of three probabilities. First is the probability that an ORF already exists, and a mutation causes transcription to emerge ($P_\textit{RNA-gain}$) but doesn't disrupt the ORF ($P_\textit{ORF-stay}$). Second is the probability that the DNA is transcribed, and an ORF emerges due to mutations ($P_\textit{ORF-gain}$) while transcription stays intact ($P_\textit{RNA-stay}$). Third is the probability that neither of the two features already exist and both emerge at the same time due to mutations (this probability is very small, and negligible). The following equation describes the probability of gene gain:

@@

Likewise, the probability of gene loss, given the gene (transcription and ORF) is already present, is the sum of the probabilities that transcription ($P_\textit{RNA-loss}$) or the ORF is lost due to mutations ($P_\textit{ORF-loss}$) due to mutations. Specifically:

@@

In previous sections, we showed that mutation bias affects the rate of \textit{de novo} gene emergence and loss. We next turned our attention to whether this bias affects the composition (and thereby the chemistry) of \textit{de novo} proteins. To this end, we first asked if the expected frequency of different amino acids is uniform, and found that it is not uniform. Amino acids like leucine (L) and serine (S) have a higher expected frequency than other amino acids. On the other hand, amino acids like methionine (M) and tryptophan (W) are less probable (\autoref{aafreq}A). This non-uniformity is to a great extent determined by number of degenerate codons for an amino acid. However, GC content also plays a role. For example, leucine, arginine and serine, all have six codons each. However, leucine is more probable than serine which in turn is more probable than arginine (\autoref{aafreq}A).

@@

We found that amino gain probabilities were also non-uniform across different amino acids (\autoref{aafreq}B) but had a good positive correlation with the expected frequency (Pearson $\rho = 0.955, P<10^{-10}$).

@@

The loss probability was also not uniformly distributed (\autoref{aafreq}C) but it did not significantly correlate with the expected frequency of the amino acids (Pearson $\rho = -0.255, P = 0.2818$).

@@

As we mentioned in previous paragraph, we restricted this analysis to substitutions that can be achieved via single nucleotide mutations. We found that most substitutions indeed had a preferred direction ($\textit{D}_{xy} \neq 0$,  \autoref{aafreq}D). The median absolute value of $\textit{D}_{xy}$ was 0.905 and the maximum value was 2.836. That is, in 38 out of 75 amino acid pairs we analysed, one direction of substitution was at least 1.87 times more probable than the other. For example, E$\to$K substitution is 1.8 times more probable than K$\to$E substitution. This may appear as a small number but many such unidirectional substitutions occurring in an evolving protein could indicate a directional selection. We also performed an analogous analysis where we defined the substitution directionality using the ratio of corresponding amino acid gain probabilities (\autoref{aagaineq}). Using this definition reduces the range of directionality values ($\textit{D}_{xy}$) such that the median value is 0.38. Although this more stringent definition reduces the intensity of the directionality, it does not change the qualitative results.

@@

\begin{align}
P_{\alpha\textit{-gain}} & = \sum_{\genfrac{}{}{0pt}{1}{x}{\ |\ \alpha_x>0}} \sum_{\genfrac{}{}{0pt}{1}{y}{\ |\ \alpha_y \geq 0}} P_y \times \mu_{y\to x}\label{alphagain}\\[1em]
%
P_{\alpha\textit{-loss}} &= \sum_{\genfrac{}{}{0pt}{1}{x}{\ |\ \alpha_x>0}} \sum_{\genfrac{}{}{0pt}{1}{y}{\ |\ \alpha_y \geq 0}} P_x \times \mu_{x\to y}\label{alphaloss}
\end{align}

@@

@@

\subsection{Probabilities of finding, gaining, and losing a DNA sequence}

\subsubsection{Probability of finding a DNA sequence}

@@

\subsection{Probabilities of finding, gaining, and losing DNA features}

@@

The probability that a DNA feature exists, is the sum of probabilities of every DNA sequence in that set.

@@

\begin{table}[!t]
\centering
\begin{tabular}{@{}ccc@{}}
\toprule
Position  & Accepted nucleotides & \begin{tabular}[c]{@{}c@{}}Number of Variants\\ (consensus + suboptimal)\end{tabular} \\ \midrule
1         & T$\gg$C$>$A,G        & $1+3$                                                                                 \\ \midrule
2         & A$\gg$T              & $1+1$                                                                                 \\ \midrule
3         & T$\gg$A,C            & $1+2$                                                                                 \\ \midrule
4         & A$\gg$T              & $1+1$                                                                                 \\ \midrule
5         & T/A                  & $2+0$                                                                                 \\ \midrule
6         & A$\gg$G$>$C/T        & $1+3$                                                                                 \\ \midrule
7         & A/T$>$G$>$C          & $2+2$                                                                                 \\ \midrule
8         & G/A$>$C/T            & $2+2$                                                                                                                                                                 \\ \bottomrule
\end{tabular}
\caption{Nucleotide residues that are tolerated at different positions in a TATA-box \citep[based on][]{tata1}}
\label{tatatab}
\end{table}

@@

\begin{equation}
P_\textit{RNA} = P_\textit{polyA} \times (P_\textit{TATA} + P_\textit{Inr})
\end{equation}

@@

\begin{equation}
P_\textit{RNA-gain} = (P_\textit{TATA-stay} + P_\textit{Inr-stay})\times P_\textit{polyA-gain} + (P_\textit{TATA-gain} + P_\textit{Inr-gain})\times P_\textit{polyA-stay}
\end{equation}

@@

\begin{align}
P_\textit{RNA-loss} & = Q_\textit{TATA-only} \times P_\textit{TATA-loss}\times(1-P_\textit{Inr-gain}) \nonumber\\[1pt]
 & \quad + Q_\textit{Inr-only}\times P_\textit{Inr-loss}\times(1-P_\textit{TATA-gain}) \nonumber \\[1pt]
 & \quad + Q_\textit{both}\times P_\textit{TATA-loss} \times P_\textit{Inr-loss} + P_\textit{polyA-loss}
\end{align}

@@

\begin{equation}
P_\textit{ORF} = P_\textit{ATG} \times P_\textit{stop} \times (1 - P_\textit{stop})^{k-2}
\end{equation}

@@

As we defined in the previous section, an ORF has three requirements (start codon, stop codon, and no premature stop codon in the sequence). Thus an ORF can emerge due to mutations via three mechanisms. In each of these mechanisms one requirement is absent whereas the other two are present.

>>>>

For all calculations we used a GC-content of 0.42 which is reasonably close to the global GC-content of human, \textit{Drosophila melanogaster}, and \textit{Saccharomyces cerevisiae} genomes. We used a mutation rate of $7.8\times10^{-9}$ mutations per nucleotide position per generation, which corresponds to spontaneous mutation rate estimated in a \textit{Drosophila melanogaster} population that was subjected to periodic population bottlenecks \cite[mutation accumulation line,][]{drosophilamutrate}. We used mutation bias data from the same study \citep{drosophilamutrate} which was in agreement with another previous study on human pseudogenes \citep{humanmutrate}. shows the values of different mutation rates (\autoref{mutbias}).

@@

If a gene is found in one species but not in sister taxa, then the most common assumption is that this gene emerged only in one specific species. We asked what is the chance that the gene was present in a common ancestor, but was lost in all lineages except one. To this end we calculated the probability of gene emergence and gene loss (Materials and Methods). Specifically, the probability of gene gain can be defined as the sum of three probabilities. First is the probability that an ORF already exists, and a mutation causes transcription to emerge ($P_\textit{RNA-gain}$, \autoref{eqrnagain}) but doesn't disrupt the ORF ($P_\textit{ORF-stay}$, \autoref{eqorfstay}). Second is the probability that the DNA is transcribed, and an ORF emerges due to mutations ($P_\textit{ORF-gain}$, \autoref{eqorfgain}) while transcription stays intact ($P_\textit{RNA-stay}$, \autoref{eqrnastay}). Third is the probability that neither of the two features already exist and both emerge at the same time due to mutations (this probability is very small, and negligible). The following equation describes the probability of gene gain:

@@

Likewise, the probability of gene loss, given the gene (transcription and ORF) is already present, is the sum of the probabilities that transcription ($P_\textit{RNA-loss}$, \autoref{eqrnaloss}) or the ORF is lost due to mutations ($P_\textit{ORF-loss}$, \autoref{eqorfloss}) due to mutations. Specifically:

@@

In previous sections, we showed that mutation bias affects the rate of \textit{de novo} gene emergence and loss. We next turned our attention to whether this bias affects the composition (and thereby the chemistry) of \textit{de novo} proteins. To this end, we first asked if the expected frequency of different amino acids is uniform, and found that it is not uniform. Amino acids like leucine (L) and serine (S) have a higher expected frequency than other amino acids. On the other hand, amino acids like methionine (M) and tryptophan (W) are less probable (\autoref{aafreq}{\color{blue}A}). This non-uniformity is to a great extent determined by number of degenerate codons for an amino acid. However, GC content also plays a role. For example, leucine, arginine and serine, all have six codons each. However, leucine is more probable than serine which in turn is more probable than arginine (\autoref{aafreq}{\color{blue}A}).

@@

We found that amino gain probabilities were also non-uniform across different amino acids (\autoref{aafreq}{\color{blue}B}) but had a good positive correlation with the expected frequency (Pearson $\rho = 0.955, P<10^{-10}$).

@@

The loss probability was also not uniformly distributed (\autoref{aafreq}{\color{blue}C}) but it did not significantly correlate with the expected frequency of the amino acids (Pearson $\rho = -0.255, P = 0.2818$).

@@

As we mentioned in previous paragraph, we restricted this analysis to substitutions that can be achieved via single nucleotide mutations. We found that most substitutions indeed had a preferred direction ($\textit{D}_{xy} \neq 0$,  \autoref{aafreq}{\color{blue}D}). The median absolute value of $\textit{D}_{xy}$ was 0.905 and the maximum value was 2.836. That is, in 38 out of 75 amino acid pairs we analysed, one direction of substitution was at least 1.87 times more probable than the other. For example, E$\to$K substitution is 1.8 times more probable than K$\to$E substitution. This may appear as a small number but many such unidirectional substitutions occurring in an evolving protein could indicate a directional selection. We also performed an analogous analysis where we defined the substitution directionality using the ratio of corresponding amino acid gain probabilities (\autoref{aagaineq}). Using this definition reduces the range of directionality values ($\textit{D}_{xy}$) such that the median value is 0.38. Although this more stringent definition reduces the intensity of the directionality, it does not change the qualitative results.

@@

\begin{align}
P_{\alpha\textit{-gain}} & = \sum_{\genfrac{}{}{0pt}{1}{x}{\,|\,\alpha_x>0}} \sum_{\genfrac{}{}{0pt}{1}{y}{\,|\,\alpha_y \geq 0}} P_y \times \mu_{y\to x}\label{alphagain}\\[1em]
%
P_{\alpha\textit{-loss}} &= \sum_{\genfrac{}{}{0pt}{1}{x}{\,|\,\alpha_x>0}} \sum_{\genfrac{}{}{0pt}{1}{y}{\,|\,\alpha_y \geq 0}} P_x \times \mu_{x\to y}\label{alphaloss}
\end{align}

@@

In this work, we addressed some fundamental questions about \textit{de novo} gene evolution. Broadly, we asked how quickly these genes can emerge and be lost, whether their birth and death has a preferred trajectory, and how the composition of their protein sequence is determined. To this end, we developed mathematical models and used them to address specific questions. These models are based on mutation probability and mutation rate bias, and represent neutral evolution where DNA sequences are not evolving under natural selection. Thus our models provide an opportunity to test the alternative hypothesis that some \textit{de novo} genes are evolving under selection.

We show that genes can be lost much more rapidly than they are gained, such that a gene can be independently lost two to three times in the expected time to gain it. Therefore, presence of a \textit{de novo} gene in one species (or population) cannot be interpreted as an evidence of gene emergence in this species.

We answer the long standing question of what is the sequence of \textit{de novo} emergence. That is, whether ORF emerges first or transcription emerges first. We show that, in the absence of selection, long ORFs are likely to emerge in existing transcripts (transcription emerges first), whereas short ORFs are likely to emerge before transcription emerges in the genomic locus.

We show that transcription loss is less likely than the loss of long ORFs, but this difference is not very large. Thus one cannot reliably infer that multiple independent transcription losses for a given \textit{de novo} gene, indicates a negative selection on the protein expression.

Finally, we show that some amino acids substitutions are more likely to occur than the others, and frequent  substitutions that are expected to occur less frequently under neutral model, can suggest directional evolution.

Like all theoretical models, our model is not without limitations. First, we assume that nucleotides are uniformly distributed according to GC content. This is not true as nucleotide composition can vary significantly across the genome. Mutation rates are also not uniform across the genome and can vary significantly \citep{mutbiasArabidopsis}. However, our model can be adapted to investigate specific genomic loci, where information on the local GC content and mutation rate can more accurately determine the \textit{de novo} gene evolution in these loci. Second, our model is based on parameters such as mutation rate and mutation bias, which we obtained from published data from humans and \textit{D.melanogaster} \citep{humanmutrate,drosophilamutrate}. The accuracy of the model will depend on the accuracy of parameters. More importantly, different species may have a different spectrum of mutations \citep{joshmutbias}, especially so for species that are highly distant from each other. Thus our inferences based on our chosen parameters will not universally apply to all species. Third, we assume that all mutations are independent. This is not the case for tandem mutations, where multinucleotide mutations occur can more frequently than their expected frequency under the assumption of mutational independence \citep{MNM}. Finally, we ignore insertions, deletions and transpositions as mechanisms of mutation. Insertions and deletions (indels) are reported to be less frequent than substitutions \citep{drosophilamutrate}, and both indels transpositions cause larger sequence alterations than point mutations. However, our modeling methodology allows incorporation of these mechanisms as a source of mutations.

As the famous saying goes, ``all models are wrong but some are useful'' \citep{GEBox}. This is true for our models too. They are not accurate but they can be useful for making predictions that can lead a more focused experimental validation. These models can be made more complex to accommodate diverse molecular mechanisms driving gene evolution. Therefore our work opens up an opportunity for theoretical biologists and computational biologists to develop models that are accurately describe their system of interest.

@@

\subsection{Probabilities of finding, gaining, and losing a DNA sequence}
\label{methbasic}

\subsubsection{Probability of finding a DNA sequence}
\label{methprob}

@@

\subsection{Probabilities of finding, gaining, and losing DNA features}
\label{methfeatures}

@@

The probability that a DNA feature exists, is the sum of probabilities of every DNA sequence in that set (\autoref{methprob}).

@@

\begin{table}[H]
\centering
\begin{tabular}{@{}ccc@{}}
\toprule
Position  & Accepted nucleotides & \begin{tabular}[c]{@{}c@{}}Number of Variants\\ (consensus + suboptimal)\end{tabular} \\ \midrule
1         & T$\gg$C$>$A,G        & $1+3$                                                                                 \\ \midrule
2         & A$\gg$T              & $1+1$                                                                                 \\ \midrule
3         & T$\gg$A,C            & $1+2$                                                                                 \\ \midrule
4         & A$\gg$T              & $1+1$                                                                                 \\ \midrule
5         & T/A                  & $2+0$                                                                                 \\ \midrule
6         & A$\gg$G$>$C/T        & $1+3$                                                                                 \\ \midrule
7         & A/T$>$G$>$C          & $2+2$                                                                                 \\ \midrule
8         & G/A$>$C/T            & $2+2$                                                                                                                                                                 \\ \bottomrule
\end{tabular}
\caption{Nucleotide residues that are tolerated at different positions in a TATA-box \citep[based on][]{tata1}}
\label{tatatab}
\end{table}

@@

\begin{equation}
P_\textit{RNA} = P_\textit{polyA} \times (P_\textit{TATA} + P_\textit{Inr})
\label{eqrnaprob}
\end{equation}

@@

\begin{equation}
P_\textit{RNA-gain} = (P_\textit{TATA-stay} + P_\textit{Inr-stay})\times P_\textit{polyA-gain} + (P_\textit{TATA-gain} + P_\textit{Inr-gain})\times P_\textit{polyA-stay}
\label{eqrnagain}
\end{equation}

@@

\begin{align}
P_\textit{RNA-loss} & = Q_\textit{TATA-only} \times P_\textit{TATA-loss}\times(1-P_\textit{Inr-gain}) \nonumber\\[1pt]
 & \quad + Q_\textit{Inr-only}\times P_\textit{Inr-loss}\times(1-P_\textit{TATA-gain}) \nonumber \\[1pt]
 & \quad + Q_\textit{both}\times P_\textit{TATA-loss} \times P_\textit{Inr-loss} + P_\textit{polyA-loss}
\label{eqrnaloss}
\end{align}

\subsubsection{Probability that transcription remains intact}

Transcription remains intact if neither the promoter, nor the poly-A signal is lost. More specifically, we define this probability ($P_\textit{RNA-stay}$) as:

\begin{equation}
P_\textit{RNA-stay} = (P_\textit{TATA-stay} + P_\textit{Inr-stay}) \times P_\textit{polyA-stay}
\label{eqrnastay}
\end{equation}

@@

\begin{equation}
P_\textit{ORF} = P_\textit{ATG} \times P_\textit{stop} \times (1 - P_\textit{stop})^{k-2}
\label{eqorfprob}
\end{equation}

@@

As we defined in the previous section, an ORF has three requirements (start codon, stop codon, and no premature stop codon in the sequence). Thus an ORF can emerge due to mutations via three mechanisms. In each of these mechanisms, one requirement is initially absent whereas the other two are present. Then mutations cause the missing requirement to emerge while not disrupting the other two requirements. More specifically, the ORF can be gained via the following three mechanisms:
\begin{enumerate}
\item Gain of a start codon ($P_\textit{ATG-gain}$) while a stop codon continues to exist at the end of the ORF ($P_\textit{stop-stay}$), and there is no emergence of stop codon within the ORF ($1- P_\textit{stop} - P_\textit{stop-gain}$).
\item Gain of a stop codon ($P_\textit{stop-gain}$), while a start codon continues to exist at the beginning of the ORF ($P_\textit{ATG-stay}$), and there is no emergence of stop codon within the ORF.
\item Loss of a premature stop codon, at any of the $k-2$ codon positions within the ORF ($P_\textit{stop-gain}$). At the same time the start codon and the stop codon remain undisrupted by mutations, and no stop codon emerges at any of the other $k-3$ positions.
\end{enumerate}

Thus we define the probability of ORF gain ($P_\textit{ORF-gain}$) as:

\begin{align}
P_\textit{ORF-gain} & = \quad P_\textit{ATG-gain}\times P_\textit{stop-stay} \times (1- P_\textit{stop} - P_\textit{stop-gain})^{k-2} \nonumber \\[1pt]
& \quad + P_\textit{ATG-stay}\times P_\textit{stop-gain} \times (1- P_\textit{stop} - P_\textit{stop-gain})^{k-2} \nonumber \\[1pt]
& \quad + P_\textit{ATG-stay}\times P_\textit{stop-stay} \times P_\textit{stop-loss}\times(k-2) \times (1- P_\textit{stop} - P_\textit{stop-gain})^{k-3} 
\label{eqorfgain}
\end{align}

\subsubsection{Probability of ORF loss}

ORF can be lost when any of its three requirements are lost. We thus define the conditional probability of ORF loss as:

\begin{equation}
P_\textit{ORF-loss} = P_\textit{ATG-loss} + P_\textit{stop-loss} + (k-2)\times \frac{P_\textit{stop-gain}}{1-P_\textit{stop}}
\label{eqorfloss}
\end{equation}

The last term in this equation describes the conditional probability of stop-gain, given the assumption that no stop codon exists within the ORF.

\subsubsection{Probability that ORF remains intact}

We assumed that an ORF of a certain length remains intact if none of the necessary features are lost. However, the ORF sequence can mutate to cause non-synonymous changes in the translated protein sequence. This condition applies to all the three ORF probabilities described above. We define the probability that an ORF remains intact ($P_\textit{ORF-stay}$) as:

\begin{equation}
P_\textit{ORF-stay} = P_\textit{ATG-stay} \times P_\textit{stop-stay} \times (1 - P_\textit{stop} - P_\textit{stop-stay})^{k-2}
\label{eqorfstay}
\end{equation}

\subsection{Probability of amino acid substitutions}

We calculated the probability of amino acid substitutions based on nucleotide substitution in the corresponding codons. In this case, multiple codons that code for the same amino acid constitute a feature set (\autoref{methfeatures}). Similarly when analysing hydrobhobic to non-hydrophobic substitutions, all codons that represent hydrophobic amino acids form a feature set (and \textit{vice versa}).

\section{Data availability}
We performed all calculations using Julia programming language, and the all scripts are freely available on GitHub: BharatRaviIyengar/DeNovoEvolution.
